Using numerical modelling to investigate temperature and wave exposure as drivers of macroalgal characteristics

### 2019-08-27 morphology_paper.Rmd
adding reference list. final output updated with citations completed thus far

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera (Bolton 2010), nevertheless form the basis of one of the most productive ecosystems globally (Mann 1973). Kelps generally have a dependence on cool, temperate and arctic seawater temperatures (Bolton 2010; @Mohring et al. 2014; Cavanaugh et al. 2011; Dayton 1985), and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and subcanopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates (Mann 1973; Steneck 2002), each depending on a wide suite of ecological services provided by the kelp forests (Gaines & nRoughgarden 1987; Paul & Steneck 1993; Levin 1994; Willis & Anderson2003; Anderson 1997; Dayton 1985). Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in the distribution (Gorman et al. 2013], abundance (Cavanaugh et al. 2011; Dayton 1998), diversity (Wing 2007; Wernberg & Goldberg 2008), composition (Leliaert et al. 2000; Dayton 1985; Norderhaug et al. 2012; Harley et al. 2012), growth (Cousens 1982) and productivity (Pedersen & Nejrup 2012; Dayton 1998; Krumhansl & Scheibling 2012) of kelps.

Temperature and wave exposure have been shown to play a role in driving kelp morphology. For example, a study by Serisawa et al. (2002) compared the morphology of Ecklonia cava Kjellman growing in warm temperate and cool temperate morphologies. The results showed that wrinkles in the blade seem to be a characteristic of warm temperate regions. However this study did not take into account the interaction with other environmental variables such as wave exposure. Wave exposure has also been shown to be the main drivers of morphological adaptation across various kelp species. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility (Denny & Gaylord 2002). For example, a study by Wernberg & Thomsen (2005) examined the consistency of wave exposure as a driver of *Ecklonia radiata* (C. Agardh) J. Agardh across a broad geographic range, and  showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites. The morphological adaptation to wave exposure must be kept in balance with other important processes, such as nutrient assimilation and light absorption, which are dependent on the amount of surface area of the blades. Therefore, there is a trade-off between reducing overall drag and maintaining nutrient and photosynthetic ability.  The morphological adapatibility of kelps is driven by a combination of environmental factors that in turn do not act independently of one another. A study by Wernberg et al. (2003) investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales (Wernberg et al. 2003). Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Wave exposure may be greater in shallower environments, or it may be reduced through damping effects of kelp further offshore and natural barriers such as rock outcrops; also, seawater temperature may be higher in shallow water environments due to higher solar irradiance.

Other species of kelp, such as *Laminaria pallida* and *Ecklonia maxima* (Osbeck) Papenfuss which are important habitat forming seaweeds that exist around the coast of South Africa, offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between deep and shallow water environments. Although both these species exist in the subtidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in from the subtidal to shallower waters. Therefore, *E. maxima* is exposed to variations in wave, swell and temperature, while *L. pallida* is exposed to variations in waves and temperature. (Molloy & Bolton 1996) investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics; when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by Rothman et al. (2017) investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. They suggested that turbidity was the environmental driver responsible for this change.

@@

Sites were chosen to represent an array of environmental gradients, as indicated in Fig. 1. St. Helena Bay and Betty's Bay constituted the north western and south eastern boundary sites, respectively. These sites are roughly 300 km apart, and lie within separate marine provinces[3] and cover a cline in wave exposure and temperature. Study sites span across the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed as a cool temperate region, which is defined as a region where mean monthly temperatures are always above 10°C and always below 15°C (Smit et al. 2013). East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone (Smit et al. 2013). The Agulhas Marine Province is characterised by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter and is classified as warm temperate region (Smit et al. 2013). Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean (Reason et al. 2006). These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter (Veitch et al. 2018). The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay (Shipley 1964). Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity.

```{r map study sites, echo=FALSE, fig.height=7, fig.width=6, message=FALSE, warning=FALSE}
library(png)
library(grid)

@@

The topography and elevation along the Cape Peninsula channel and shield winds along False Bay. This is however absent in winter, where strong northerly winds are prevalent from St. Helena Bay to Betty's Bay (Field et al. 1980; Jury et al. 1985; Andrews & Hutchings 1980; Jury 1980). Patterns are also evident in the wave data. Seasonal variations in significant wave height (Hs), peak period and wave direction are also present. The direction of the dominant swell changes to the south west in winter, generated by strong low pressures that originate from the southern ocean (Reason et al. 2006), which False Bay is sheltered from (Shipley 1964; Atkins 1970; Dufois & Rouault 2012). In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an increased variability of Hs and Tp in this region. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world (Norderhaug et al. 2012; Leliaert et al. 2000). It is due to the near consistent south-westerly swell and the complex orography around the peninsula that the wave energy distribution around the Cape Peninsula varies significantly over a small geographical area.

@@

Shallow water temperatures were sourced from The South African Coastal Temperature Network (SACTN) website (https://github.com/ajsmit/SACTN). In terms of nearshore temperature, *in situ* data are preferred over satellite SST, which have shown to exhibit large biases (Smit et al. 2013). Linear interpolated SST were calculated for sites where *in situ* recorders were absent. These data were used to group sites into 'Cool temperate' and 'Warm temperate' categorisations to investigate the possible affect of thermal regime on kelp morphological characteristics.

@@

Wave model variable data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs (DEA) and produced by the Council for Scientific and Industrial Research (CSIR) (Rautenbach 2015). The South African wave climate was modeled via 20 spectral numerical wave models that simulated the offshore wave climate to the nearshore. The boundary conditions of these models were obtained by using the NOAA Wave Watch III (WWIII) model output, distributed via the National Centers for Environmental Prediction (NCEP) product (Office of the Director 2000; Environmental Modeling Center / Marine Modeling Branch 2005). The particular hindcast product utilised during the DEA-CSIR study spans 1994-2013 at a 3-hour resolution. These data were then used to model swell propagation into the coastal models while wind waves (seas) were generated via stationary computations in the Simulating Waves in the Nearshore SWAN model (Booij et al. 1997). The assumption of stationary computations are acceptable as the model domains were small enough so the temporal variation of the model boundary were slower than the time it takes for that boundary condition to propagate to the coast. SWAN allows one to extract wave variables from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modelled and output produced at both the 7 meter and 15 meter isobaths. A 200 meter resolution was used as the False Bay computational grid was nested within a larger grid (1 kilometer resolution). This allowed for a computational effective wave resolution of increasing resolution, from the NCEP, low resolution output to nested, high resolution coastal output. For Table Bay and east of Cape Hangklip the resolution was 500 meters and also had output at the 7 meters and 15 meter isobaths. These contour outputs were chosen in the original study by the CSIR as most engineering run-up calculations require wave parameter information at these contour depths and were the main focus of the original study. For this study the 7 meter contours were used. These data were then used to calculate over all wave power (kW/m) as this was considered the best measure of overall wave exposure. Annual wave power was plotted and categorised into different wave exposure categories which ranged from fully sheltered to extremely exposed.

@@

The morphological characteristics of both species are presented in Fig. 2. *Laminaria pallida* is characterised by a single smooth blade which is divided longitudinally into sections, and develops from a single meristematic region located at the junction between the blade and the stipe (Dyer 2018). This species has a solid stipe but develops a hollow stipe along the west coast northward and into Namibia (Rothman et al. 2017). *Ecklonia maxima* consists of a single primary blade which develops above a gas-filled float and a hollow stipe below. Secondary blades are produced laterally from the primary blade from several meristematic regions along the margins of the primary blade, known as digits (Dyer 2018). Both species are held to the substrate by finger-like haptera, collectively known as the holdfast.

```{r schematic, echo=FALSE, fig.cap='Schematic of *L. pallida* and *E. maxima* morphology.', fig.height=7, fig.width=7}

@@

Characterisation of morphological characteristics in relation to temperature and wave exposure non-parametric statistical methods were used. Morphological characteristics were categorised according to temperature regime and wave exposure category for each each species and population of kelp in this study. Significant differences between temperatre regime and wave exposure categories were tested with wilcox tests. Notched boxplots allowed comparison between individual categories of wave exposure. In order to investigate specific drivers of morphological characteristics a distance-based redundancy analysis was performed using the *rda* function in the *vegan* software package (Oksanen et al. 2016) in R (R Core Team, 2019). The abiotic data covered all study sites for both species of kelp. Both morphology and abiotic data (temperature, waves and wind) were standardised using the *decostand* function in the *vegan* software package; morphology data were used as response variables and the abiotic data as explanatory variables. To determine the explanatory variables that best describe patterns in the response data, a full RDA was performed using a complete set of explanatory variables. Forward selection was then used to reduce the number of explanatory variables as well as prevent the inflation of overall type I error. To further improve the model, pairwise coefficients and Variance Inflation Factor (VIF) were calculated to identify variables with high multicollinearity. The computation of the parsimonious RDA was followed by permutation tests of the adjusted *R*2 to assess significance of constraints.

@@

```{r rda lam plot, echo=FALSE, fig.height=4, fig.width=4, message=FALSE, warning=FALSE}
lam_rda
```

@@

```{r rda eck plots, echo=FALSE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE}
ggarrange(rda_eck_deep, rda_eck_shallow, labels = "AUTO", common.legend = TRUE, legend = "bottom", ncol = 1, nrow = 2)
```
Forward selection, assessment of VIF and Pearson correlation coefficients allowed us to retian the most parsimonious descriptors of shallow *E. maxima* morphological characteristics with an adjusted *R*2 of 0.36, explaining 61% of the variation (global permutation test on final model: d.f = 4, *F* = 2.41, *p* = 0.013). The model consisted of one significant axis, RDA1 (p = 0.02) and explained 62% of the variation. Stipe length, stipe mass and stipe circumference were positivley influenced by *max_temp* and *mean_temp* along RDA1, and explained 50% and 55% of the variation respectivley. The remaining morphological characteristics were positivley influenced by by *tp_max* and negativley inlfuenced by *dirw_median* along RDA2 and explained 86% and 94% of the variation respectivley. There was overlapping of clusters according to region with no clear patterns in sites evident.

@@

This study considered important drivers of kelp morphology that have been identified in previous research (Fowler-Walker et al. 2006; Miller et al. 2011; Fowler-Walker et al. 2005; Stewart et al. 2009; Wing et al. 2007; Molloy & Bolton 1996; Pedersen & Nejrup 2012; Wernberg & Thomsen 2005), namely temperature and wave exposure, around the Cape Peninsula on two prolific kelp species. Unlike previous research, this study used a numerical model to provide the various variables of wave exposure. We also considered wind in addition to temperature and wave variables, since wind is an important component of surface driven waves. The results show that specific variables of wave exposure are the main drivers of kelp morphology with temperature and wind playing only minor roles. Furthermore, the investigation of differences in morphological characteristics between shallow and deep populations of *E. maxima * suggests that in low wave exposure environments the role of temperature as a morphological driver increases, such as in False Bay.

The complex geomorphology of the western Cape coastline creates an ideal natural laboratory for studies of kelps in interaction with their environment. The annual and seasonal patterns in both wind influence the thermal regime around the coastline by inducing upwelling. Upwelling brings deep, cool nutrient rich water to coastal areas and therefore induces a decrease in temperature (Field et al. 1980; Andrews & Hutchings 1980; Gill & Clarke 1974; Cram 1970; Rouault et al. 2010; Blanke et al. 2002. The wind speed and wind direction help drive upwelling in summer on the western side of the Cape peninsula, where southerly winds blow parallel to the coast and trigger upwelling (Field et al. 1980; Rouault et al. 2010). This is not the case for False Bay which is shielded from the dominating wind and swell the Cape Peninsula is subject to, with the exception being the winter months where upwelling becomes more favourable for the region due to the change in swell and wind direction, causing lower temperatures to dominate during this time (Rouault et al. 2010; Andrews & Hutchings 1980). Increased wind speed at sites along the west side of the peninsula in a southerly direction trigger upwelling events along the western side of the peninsula (Rouault et al. 2010; Field et al. 1980). We however see a possible lack of upwelling events for sites such as Betty’s Bay and Batsata Rock.

>>>>

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], nevertheless form the basis of one of the most productive ecosystems globally [@Mann1973]. Kelps generally have a dependence on cool, temperate and arctic seawater temperatures [@Bolton2010; @Mohring2014-bk; @Cavanaugh2011-ib; @Dayton1985], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and subcanopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Steneck2002], each depending on a wide suite of ecological services provided by the kelp forests [@Gaines1987-bg; @Paul1993-bm; @Levin1994-qf; @Willis2003-xo; @anderson1997; @Dayton1985]. Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in the distribution [@Gorman2013-nj], abundance [@Cavanaugh2011-ib; @Dayton1998-xw], diversity [@Wing2007-tv; @Wernberg2008], composition [@Leliaert2000a; @Dayton1985; @Norderhaug2012-ks; @Harley2012], growth [@Cousens1982-ck] and productivity [@Pedersen2012-vr; @Dayton1998-xw; @Krumhansl2012] of kelps.

Temperature and wave exposure have been shown to play a role in driving kelp morphology. For example, a study by @Serisawa2002-dc compared the morphology of Ecklonia cava Kjellman growing in warm temperate and cool temperate morphologies. The results showed that wrinkles in the blade seem to be a characteristic of warm temperate regions. However this study did not take into account the interaction with other environmental variables such as wave exposure. Wave exposure has also been shown to be the main drivers of morphological adaptation across various kelp species. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002-ob]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *Ecklonia radiata* (C. Agardh) J. Agardh across a broad geographic range, and  showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites. The morphological adaptation to wave exposure must be kept in balance with other important processes, such as nutrient assimilation and light absorption, which are dependent on the amount of surface area of the blades. Therefore, there is a trade-off between reducing overall drag and maintaining nutrient and photosynthetic ability.  The morphological adapatibility of kelps is driven by a combination of environmental factors that in turn do not act independently of one another. A study by @Wernberg2003-mp investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp]. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Wave exposure may be greater in shallower environments, or it may be reduced through damping effects of kelp further offshore and natural barriers such as rock outcrops; also, seawater temperature may be higher in shallow water environments due to higher solar irradiance.

Other species of kelp, such as *Laminaria pallida* and *Ecklonia maxima* (Osbeck) Papenfuss which are important habitat forming seaweeds that exist around the coast of South Africa, offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between deep and shallow water environments. Although both these species exist in the subtidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in from the subtidal to shallower waters. Therefore, *E. maxima* is exposed to variations in wave, swell and temperature, while *L. pallida* is exposed to variations in waves and temperature. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics; when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. They suggested that turbidity was the environmental driver responsible for this change.

@@

Sites were chosen to represent an array of environmental gradients, as indicated in Fig. 1. St. Helena Bay and Betty's Bay constituted the north western and south eastern boundary sites, respectively. These sites are roughly 300 km apart, and lie within separate marine provinces and cover a cline in wave exposure and temperature. Study sites span across the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed as a cool temperate region, which is defined as a region where mean monthly temperatures are always above 10°C and always below 15°C [@Smit2013a]. East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone [@Smit2013a]. The Agulhas Marine Province is characterised by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter and is classified as warm temperate region [@Smit2013a]. Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean [@Reason2006]. These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter [@Veitch2019]. The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay [@Shipley1964]. Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity.

```{r map study sites, echo=FALSE, fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
library(png)
library(grid)

@@

The topography and elevation along the Cape Peninsula channel and shield winds along False Bay. This is however absent in winter, where strong northerly winds are prevalent from St. Helena Bay to Betty's Bay [@Field1980a; @Jury1985-nr; @Andrews1980; @Jury1980-ov]. Patterns are also evident in the wave data. Seasonal variations in significant wave height (Hs), peak period and wave direction are also present. The direction of the dominant swell changes to the south west in winter, generated by strong low pressures that originate from the southern ocean [@Reason2006], which False Bay is sheltered from [@Shipley1964; @Atkins1970-bx; @Dufois2012-vy]. In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an increased variability of Hs and Tp in this region. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world [@Norderhaug2012-ks; @Leliaert2000a]. It is due to the near consistent south-westerly swell and the complex orography around the peninsula that the wave energy distribution around the Cape Peninsula varies significantly over a small geographical area.

@@

Shallow water temperatures were sourced from The South African Coastal Temperature Network (SACTN) website (https://github.com/ajsmit/SACTN). In terms of nearshore temperature, *in situ* data are preferred over satellite SST, which have shown to exhibit large biases [@Smit2013a]. Linear interpolated SST were calculated for sites where *in situ* recorders were absent. These data were used to group sites into 'Cool temperate' and 'Warm temperate' categorisations to investigate the possible affect of thermal regime on kelp morphological characteristics.

@@

Wave model variable data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs (DEA) and produced by the Council for Scientific and Industrial Research (CSIR) [@Rautenbach2015a]. The South African wave climate was modeled via 20 spectral numerical wave models that simulated the offshore wave climate to the nearshore. The boundary conditions of these models were obtained by using the NOAA Wave Watch III (WWIII) model output, distributed via the National Centers for Environmental Prediction (NCEP) product (Office of the Director 2000; Environmental Modeling Center / Marine Modeling Branch 2005). The particular hindcast product utilised during the DEA-CSIR study spans 1994-2013 at a 3-hour resolution. These data were then used to model swell propagation into the coastal models while wind waves (seas) were generated via stationary computations in the Simulating Waves in the Nearshore SWAN model [@Booij1997-nx]. The assumption of stationary computations are acceptable as the model domains were small enough so the temporal variation of the model boundary were slower than the time it takes for that boundary condition to propagate to the coast. SWAN allows one to extract wave variables from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modelled and output produced at both the 7 meter and 15 meter isobaths. A 200 meter resolution was used as the False Bay computational grid was nested within a larger grid (1 kilometer resolution). This allowed for a computational effective wave resolution of increasing resolution, from the NCEP, low resolution output to nested, high resolution coastal output. For Table Bay and east of Cape Hangklip the resolution was 500 meters and also had output at the 7 meters and 15 meter isobaths. These contour outputs were chosen in the original study by the CSIR as most engineering run-up calculations require wave parameter information at these contour depths and were the main focus of the original study. For this study the 7 meter contours were used. These data were then used to calculate over all wave power (kW/m) as this was considered the best measure of overall wave exposure. Annual wave power was plotted and categorised into different wave exposure categories which ranged from fully sheltered to extremely exposed.

@@

The morphological characteristics of both species are presented in Fig. 2. *Laminaria pallida* is characterised by a single smooth blade which is divided longitudinally into sections, and develops from a single meristematic region located at the junction between the blade and the stipe [@Dyer2018-rv]. This species has a solid stipe but develops a hollow stipe along the west coast northward and into Namibia [@Rothman2017-pn]. *Ecklonia maxima* consists of a single primary blade which develops above a gas-filled float and a hollow stipe below. Secondary blades are produced laterally from the primary blade from several meristematic regions along the margins of the primary blade, known as digits [@Dyer2018-rv]. Both species are held to the substrate by finger-like haptera, collectively known as the holdfast.

```{r schematic, echo=FALSE, fig.cap='Schematic of *L. pallida* and *E. maxima* morphology.', fig.height=6, fig.width=6}

@@

Characterisation of morphological characteristics in relation to temperature and wave exposure non-parametric statistical methods were used. Morphological characteristics were categorised according to temperature regime and wave exposure category for each each species and population of kelp in this study. Significant differences between temperatre regime and wave exposure categories were tested with wilcox tests. Notched boxplots allowed comparison between individual categories of wave exposure. In order to investigate specific drivers of morphological characteristics a distance-based redundancy analysis was performed using the *rda* function in the *vegan* software package [@Oksanen2015-vj] in R (R Core Team, 2019). The abiotic data covered all study sites for both species of kelp. Both morphology and abiotic data (temperature, waves and wind) were standardised using the *decostand* function in the *vegan* software package; morphology data were used as response variables and the abiotic data as explanatory variables. To determine the explanatory variables that best describe patterns in the response data, a full RDA was performed using a complete set of explanatory variables. Forward selection was then used to reduce the number of explanatory variables as well as prevent the inflation of overall type I error. To further improve the model, pairwise coefficients and Variance Inflation Factor (VIF) were calculated to identify variables with high multicollinearity. The computation of the parsimonious RDA was followed by permutation tests of the adjusted *R*2 to assess significance of constraints.

@@

@@

```{r rda eck plots, echo=FALSE, fig.height=7, fig.width=9, message=FALSE, warning=FALSE}
ggarrange(lam_rda, rda_eck_deep, rda_eck_shallow, common.legend = TRUE, legend = "bottom", labels = "AUTO",ncol = 2, nrow = 2)
```

Forward selection, assessment of VIF and Pearson correlation coefficients allowed us to retian the most parsimonious descriptors of shallow *E. maxima* morphological characteristics with an adjusted *R*2 of 0.36, explaining 61% of the variation (global permutation test on final model: d.f = 4, *F* = 2.41, *p* = 0.013). The model consisted of one significant axis, RDA1 (p = 0.02) and explained 62% of the variation. Stipe length, stipe mass and stipe circumference were positivley influenced by *max_temp* and *mean_temp* along RDA1, and explained 50% and 55% of the variation respectivley. The remaining morphological characteristics were positivley influenced by by *tp_max* and negativley inlfuenced by *dirw_median* along RDA2 and explained 86% and 94% of the variation respectivley. There was overlapping of clusters according to region with no clear patterns in sites evident.

@@

This study considered important drivers of kelp morphology that have been identified in previous research [@FowlerWalker2006; @Miller2011-pm; @Fowler-Walker2005; @Stewart2009; @Wing2007-tv; @Molloy1996-la; @Pedersen2012-vr; @Wernberg2005c], namely temperature and wave exposure, around the Cape Peninsula on two prolific kelp species. Unlike previous research, this study used a numerical model to provide the various variables of wave exposure. We also considered wind in addition to temperature and wave variables, since wind is an important component of surface driven waves. The results show that specific variables of wave exposure are the main drivers of kelp morphology with temperature and wind playing only minor roles. Furthermore, the investigation of differences in morphological characteristics between shallow and deep populations of *E. maxima * suggests that in low wave exposure environments the role of temperature as a morphological driver increases, such as in False Bay.

The complex geomorphology of the western Cape coastline creates an ideal natural laboratory for studies of kelps in interaction with their environment. The annual and seasonal patterns in both wind influence the thermal regime around the coastline by inducing upwelling. Upwelling brings deep, cool nutrient rich water to coastal areas and therefore induces a decrease in temperature [@Field1980a; @Andrews1980; @Gill1974-nt; @Cram1970-jw; @Rouault2010-gi; @Blanke2002]. The wind speed and wind direction help drive upwelling in summer on the western side of the Cape peninsula, where southerly winds blow parallel to the coast and trigger upwelling [@Field1980a; @Rouault2010-gi]. This is not the case for False Bay which is shielded from the dominating wind and swell the Cape Peninsula is subject to, with the exception being the winter months where upwelling becomes more favourable for the region due to the change in swell and wind direction, causing lower temperatures to dominate during this time [@Rouault2010-gi; @Andrews1980]. Increased wind speed at sites along the west side of the peninsula in a southerly direction trigger upwelling events along the western side of the peninsula [@Rouault2010-gi; @Field1980a]. We however see a possible lack of upwelling events for sites such as Betty’s Bay and Batsata Rock.

We provide strong support that variations in the environmental variables, particularly wave exposure variables, are driving kelp morphological characteristics around the Cape Peninsula. Morphological adaptation due to water motion may manifest in a number of ways in high wave energy environments. For instance, reduction of blade thickness, blade elongation, increase of stipe length, increase in stipe circumference and force of attachment ([@FowlerWalker2006; @Wernberg2005c; @Friedland1995-vi; @Denny2002; @Bekkby2014; @Denny1997; Denny2002-ob]) have been suggested in previous studies. Although this study did not measure force of attachment, other morphological responses to temperature and wave variables were evident. Species-specific responses are evident to both wave exposure and temperature. In cool-temperate environments *L. pallida* tended to show increases in certain morphological characteristics (stipe mass, stipe length and stipe diameter) while in the warm-temperate environments these were significantly lower. This was also true for deep *E. maxima* populations, which were significantly taller and thinner for kelp in the warm-temperate compared to kelp in the cool-temperate. This has been documented previously by ?????, and show that the 'slender' kelp morphology is a response to warmer temperatures. In warmer environments kelp metabolism increases, using more nutrients than under cooler conditions causing the kelp to become slender **more needed here**. It should be noted however, that from these analyses that in general the cool-temperate region is more exposed to waves than that of the warm-temperate region. Therefore, the significantly larger morphological characteristics for kelp from the cool-temperate region may overlap with responses to wave exposure which was exhibited strongly in both species. When grouped by wave exposure category, *L. pallida* characteristics showed a significant increase in length for the 'Extremely sheltered' category (6-10 kW/m; figure ??) which maybe a 'go-with-the-flow' tactic. Kelp are able to increase flexibility by increasing stipe length which increases the extension capabilities of kelp to a passing wave. However, increasing stipe length is only beneficial under lower exposure levels as a long stipe actually increases overall drag on the plant under higher exposure levels. This is reflected in Figure ?? which shows an overall reduction of *L pallida* morphological characteristics suggesting a size reducing tactic to cope with higher levels of wave exposure. The morphological characteristics of *E. maxima* show similar patterns when grouped to wave exposure. The results suggest that deep *E. maxima* populations exhibit a different response as well as a higher wave exposure threshold. Unlike *L. pallida*, *E. maxima* exhibits a size reducing tactic with in an increase in exposure level by significantly decreasing stipe mass, stipe length, stipe circumference, total length and frond length. This size reduction tactic has been shown before (citations ???) and is regarded as strategy to reduce overall drag on the plant. However, *E. maxima* morphological characteristics changed significantly as wave exposure increases. When grouped to the higher wave exposure levels, *E. maxima* morphology exhibited a strength and flexibility increasing trait (see figure ??). Furthermore, the morphological response of *E. maxima* to wave exposure only occurs at the 'Moderately exposed' (20-30 kW/m) level compared to *L. pallida* which exhibits a response at a lower wave exposure level. This suggests that *E. maxima* can tolerate higher exposure levels before having to exhibit a morphological response. The redundancy analysis performed confirms the patterns and responses observed as well as identifying specific temperature and wave variables as drivers of kelp morphology. Kelp morphology characteristics are largely wave driven for both species but differ in terms of specific temperature and wave metrics. For instance, both species morphological characteristics are determined by swell direction (dir_median), standard deviation in peak period (tp_sd) and mean significant swell height (hs_mean), but *L. pallida* temperature driver is mean temperature (mean_temp) and *E. maxima* by minimum temperature (min_temp). Minimum temperature as an important metric for *E. maxima* may be related to nutrient availability. Low temperatures are often associated with upwelling events which bring cool, nutrient rich water into the nearshore. False Bay has comparativley little upwelling events compared to the western side of the peninsula, and so nutrients may be a limiting factor for *E. maxima* populations within False Bay (see figure ??). The kelp *L. pallida* is a sub-canopy species where more efficent mixing of water within the water column occurs (ref ??). The difference in canopy type between the species may also be the reason why *E. maxima* is driven by multiple wave metrics compared to *L. pallida*. Since *E. maxima* is a canopy-kelp, it is will be exposed to all components of a wave compared to *L. pallida* which occurs deeper in the water column.

Annual maximum and winter mean temperatures influence stipe length and frond length in shallow *E. maxima* individuals located at sites within False Bay. This was not the case for deeper *E. maxima* individuals as well as *L. pallida* which were largely driven by wave variables. The shallow *E. maxima* individuals may not experience the same amount of wave energy as do deeper populations. This may be due to wave dampening from deeper populations of *E. maxima*, which attenuate the wave energy entering coastal environments [@Kobayashi1993-os; @Jackson1984; @Mork1996-md]. The morphological variations in characteristics for deep *E. maxima* populations are not seen for shallow *E. maxima* populations. The lack of variation in shallow *E. maxima* populations was also found in previous research by [@Rothman2017-pn] who showed that light was a significant driver. This influence of light may suggest photosynthetic ability and nutrient uptake as important drivers in shallow environments. Wave exposure may be reduced in shallow environments due to the wave damping effect of deep kelp and seaweed populations [@Kobayashi1993-os; @Mork1996-md; @Jackson1984], or rocky outcrops, thereby reducing the need to adapt morphological characteristics to reduce drag. Furthermore, the reduction in differences between deep and shallow *E. maxima *at wave exposed sites on the western of the peninsula compared to sites in False Bay provides further evidence that wave exposure is not an important driver of shallow *E. maxima *populations. Increased temperatures corresponded with a reduction in  stipe circumference, frond length, stipe length, total length and stipe mass resulting in slim or ‘skinny’ kelp individuals, which supports findings from previous research by [@Mabin2013-zs; @Serisawa2002-dc].
